---
title: "Family Feud - NYT Crossword Edition"
author: "Simon Heitin"
date: "2022-09-28"
format: 
  html:
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
  message: false
  freeze: auto
categories: [tidytuesday]
---

If there's one thing that unites my family, it's a love of food[^1]. If there are two things that unite my family, they are a love of food and a love of puzzles. So, when I saw that [tidytuesday](https://github.com/rfordatascience/tidytuesday) had wrangled [a set of crossword clues and answers](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-04-19/readme.md), I knew what I had to do.

[^1]: this post is not about about food but maybe the next one should be

I emailed my family members and asked them to submit one guess as to what they thought was the most common 4-letter crossword answer.

```{r pull_data_and_tidy}
library(tidyverse)
library(lubridate)

path <- 
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-04-19/times.csv'
  
times <- read_csv(path) 

times_clean <- times %>% 
  mutate(answer = str_remove_all(answer, " "),
         answer_len = nchar(answer)) %>% 
  filter(!is.na(puzzle_date)) 

#these are the guesses. I provided a pseudonym for each family member to keep 
# things friendly. https://babynames.com/names/random-name-generator.php was a
#  big help
fam_responses_anon <- data.frame(
  stringsAsFactors = FALSE,
               answer = c("ANTE","ERRS","HOME",
                          "IDEA","DERN","BEAU","SETS","EONS","ERAS","TIME",
                          "WELL","WHAT","ACRE","STAG","ARIL", "LAST", "CLAM", "FOUR"),
            pseudonym = c("Keona","Gordian","Zai",
                          "Makaila","Olof","Quintana","Selene","Ceron",
                          "Harrison","Mignonette","Tua","Alysha","Ziva",
                          "Maddox","Mervin", "Ultimate Warrior", "Janet", "Tarzan")
   )

# get the number of occurrences for each 4-letter word
answers_ordered <- times_clean %>%
  filter(answer_len == 4) %>%
  count(answer, name = "n_appearances") %>%
  arrange(desc(n_appearances)) 

# to calculate percentiles, we need to 
answers_pctile <- answers_ordered %>%
  count(n_appearances, name = "n_answers") %>%
  arrange(desc(n_appearances)) %>%
  mutate(num_better_answers = lag(cumsum(n_answers), default = 0),
         pct_better_answers = num_better_answers / sum(n_answers),
         pctile = 1-pct_better_answers) %>%
  right_join(answers_ordered, by = "n_appearances") 

# now just get the answers from my family
family_answers_pctile <- answers_pctile %>%
  right_join(fam_responses_anon, by = "answer") %>%
  mutate(pctile = replace_na(pctile, 0),
         num_better_answers = replace_na(num_better_answers, 2000))
```

## Bum Bum Ba Da Buuuuum - Results

The plot below shows the results of our challenge. The y-axis is the percentile for how common each 4-letter word is. So, if a word is in the 90th percentile, then it is more common than 90% of all 4-letter words (that appeared in at least one crossword puzzle).

```{r plot_results}
#| fig-width: 4
#| fig-height: 6
#| fig-cap: "Family Guesses and Their Resulting Percentiles"
#| fig-cap-location: bottom
#| fig-alt: 
#|   a scatterplot with a single vertical line. The x-value is 0 for each 
#|   point and the y value corresponds to the percentile of each word. Each word 
#|   itself is labeled. The top two words are 'WELL' and 'STAG'.

library(ggrepel)
library(scales)

 family_answers_pctile %>%
  ggplot(aes(y = pctile)) +
  geom_vline(xintercept = 0) +
  geom_point(aes(x = 0), alpha = .3, stroke = 0) +
  # stat_ecdf() +
  geom_label_repel(aes(label = answer, x = 0), max.overlaps = 20, fill = "light blue") +
  scale_y_continuous(labels = label_percent(), limits = c(0,1)) +
  theme_light() +
  theme(axis.text.x=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())  +
  labs(y = "Answer Percentile")
```

A keen eye will note that the top answer, "WELL", is at the 100th Percentile. That's right, one family member chose the absolute most common 4-letter crossword clue! It's no wonder that this is also the person who made a living out of playing poker through serious study and computer simulation.

## Try For Yourself

Type an answer in ALL CAPS in the text box in the first column below. This is filtered to 4-letter answers.

```{r reactable}
#| fig-caption: 

library(reactable)
answers_pctile %>%
  select(answer, n_appearances, pctile) %>%
  mutate(percentile = scales::percent(pctile, accuracy = .01)) %>%
  select(`Answer` = answer, `Answer Percentile` = percentile, `No. of Puzzles in Which Answer Appeared` = n_appearances) %>%
  reactable::reactable(filterable = TRUE, defaultPageSize = 1)
```
